Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-10-27 TEBenchmarkPaper/article.tex
More work

\subsection{Preparation}

@@

Some problem sets are more amenable for this purpose than others. The most
suitable are those which meet the following criteria:

\begin{itemize}
\item For each problem, there should be a clear separation between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like SMTLIB, where many
  problems \emph{declare} a function, but leave its behaviour \emph{implicit} in
  the logical structure of the theorem statement.
\item Definitions should be \emph{strongly typed}, such that they can be
  translated into the native language of each ATE system (in this work, those
  languages are Haskell and Isabelle).
\item The problem set should be relevant to the desired domain. In our case, we
  desire higher-order functions and inductively defined types, which rules out
  first-order languages/logics.
\item Since it will act as our ground truth, the problem set should ideally
  contain \emph{every} ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
  definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

Once a theorem proving problem set has been chosen, we must separate the
definitions referenced by the theorems from the theorem statements themselves.
The former will be used as input to the ATE systems under evaluation, whilst the
latter form a ground truth corpus against which to compare the ATE output.

@@

We form the ground truth as above: selecting all applicable theorems
from the corpus, which is now guaranteed to be non-empty.

@@

In our experiments we have found that memory usage is also important, but rather
than complicating our analysis with an extra dimension, we instead set a limit
(as large as practical) and terminate program runs which cross it.~\footnote{We
  could treat the hardware's RAM capacity as our limit, allow the operating
  system to swap any overflow to disk, and wait for the inevitable slowdown to
  trigger the timeout. This requires no work to implement, but wastes a lot of
  time waiting for runs which could otherwise be terminated early.}

This approach process produces a single runtime, precision and recall value for
each sample. We propose two methods for analysing this data: \emph{summarising}
the performance of an ATE system, or \emph{comparing} the performance of
\emph{two} ATE systems.

@@

 - Giving each run a different sample is important for getting independent
   values which test the generality of the system. This is preferable to e.g.
   timing each sample multiple times.

 - What sorts of aggregate data to report: median vs mean, variance, quartiles,
   etc.

@@

To encourage competition in the field, it is important to compare the relative
performance of different ATE systems on the same task. Since the aggregate
statistics in the above summaries do not include details of specific runs, any
comparison based on them would have very low statistical power.

To produce more salient results, we should compare the data points directly, for
example with a \emph{paired difference test}: for each individual sample, we
find the \emph{difference} between the two systems' data \emph{for that sample}.
We then aggregate these differences, to determine whether one system performs
significantly better or worse than the other.

 - Good because: handles the variance of results well
 - Good because: approximates a normal distribution, making meaningful statistics easier
 - Good because: conservative in the face of timeouts

 - Question: do we use relative difference, e.g. [ 1.0:1.2, 1.0:1.5, ...], or do
   we use absolute differences, e.g. [100s:120s, 1s:1.5s, ...]?

@@

We call the Theory Exploration Benchmark (TEB).

@@

% evaluation

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy theory
exploration systems. To assess the scalability of their algorithms, we sampled
theories of various sizes: from theories containing only a single definition, up
to theories containing 20 definitions.

>>>>

\subsection{Preparation}
\label{section:prep}

@@

Some problem sets are more amenable for this purpose than others. The most
suitable are those which have the following properties:

\begin{itemize}
\item For each problem, there should be a clear distinction between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like SMTLIB, where many
  problems \emph{declare} a function, but leave its behaviour \emph{implicit} in
  the logical structure of the theorem statement.
\item Definitions should be \emph{strongly typed}, such that they can be
  translated into the native language of each ATE system (in this work, those
  languages are Haskell and Isabelle).
\item The problem set should be relevant to the desired domain. In our case, we
  desire higher-order functions and inductively defined types, which rules out
  first-order languages/logics.
\item Since it will act as our ground truth, the problem set should ideally
  contain every ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

Once such a problem set has been chosen, we must separate the definitions
referenced by the theorems from the theorem statements themselves. The former
will be used as input to the ATE systems under evaluation, whilst the latter
form a ground truth corpus against which to compare the ATE output.

@@

When we form the ground truth for such samples, we're guaranteed to find at
least one applicable theorem from the corpus (the one we chose as our starting
point).

@@

In our experiments we have found that memory usage is also an important part of
a system's performance, but rather than complicating our analysis with an extra
dimension, we instead set a limit (as large as practical) and terminate program
runs which cross it.~\footnote{We could treat the hardware's RAM capacity as our
  limit, allow the operating system to swap any overflow to disk, and wait for
  the inevitable slowdown to trigger the timeout. This requires no work to
  implement, but wastes a lot of time waiting for runs which could otherwise be
  terminated early.} This is in line with the expected usage of these systems:
either there is enough memory, or there isn't; implementations shouldn't be
penalised for making use of available resources.

This approach produces a single runtime, precision and recall value for each
sample. We propose two methods for analysing this data: \emph{summarising}
the performance of an ATE system, or \emph{comparing} the performance of
\emph{two} ATE systems.

@@

Each sample is only explored once, so that we cover as many independent samples
as possible to better estimate how a system's performance generalises to unseen
inputs. To combine these data into an aggregate summary depends on what we are
interested in measuring.

One general question we might ask is how a system's performance scales with
respect to the input size (the number of definitions in the theory). This is
straightforward to measure by varying the sample size, but we need some way to
combine the results from samples of the same size.

The precision and recall for multiple samples $S$ can be combined in two ways:
finding their mean value (the \emph{average of the ratios}) or by summing the
output and ground-truth counts \emph{before} taking their ratios (the
\emph{ratio of the averages}). The former estimates the precision and recall we
can expect from any \emph{particular} run, which is of direct relevance to users
of these tools. The latter estimates the proportion of discoveries that a tool
can expect to find out of 'all possible', i.e. assigning higher weight to
performance in theories which have more conjectures available to be discovered;
whilst this is of theoretical interest, it is a more difficult claim to support
empirically, so we leave its investigation to future work.

We summarise the runtimes by choosing their median, as it is more robust against
long-running outliers, and hence represents performance for a ``typical''
theory.

Another advantage to our statistical approach is that we can compute the
\emph{variance} of our data, which is unknown in the one-off evaluations which
have been performed previously. In particular, we might be concerned that such
previous evaluations used ``popular'' theories, like those involving natural
numbers, which may be unrepresentative of typical usage: any number theorist
knows how amenable natural numbers are to conjecture and exploration, and we
should not expect a typical user's theory (e.g. a software library) to be
similarly rich.

@@

To encourage competition in the field, it is important to compare the relative
performance of different ATE systems on the same task. Since the aggregate
statistics in the above summaries do not include details of specific runs, any
comparison based on them would have very low statistical power, and therefore it
is appropriate to use an alternative approach when comparing.

We propose comparing the data points directly, using a \emph{paired difference
  test}: for each individual sample, we find the \emph{difference} between the
two systems' data \emph{for that sample}. We then aggregate these
\emph{differences}, to determine whether one system performs significantly
better or worse than the other.

This neatly avoids the variance inherent in the summary statistics, and (with
enough samples) even approaches a normal distribution. Paired difference tests
are also robust to the censoring caused by timeouts: upper-bounding the run time
gives an upper bound to the (observable) differences, biasing our inference
towards the null hypothesis (indistinguishable performance) rather than
unfounded conclusions.

@@

We have applied our benchmarking methodology to the Tons of Inductive Problems
(TIP) problem set for automated theorem provers. TIP satisfies the desiderata of
section\ref{section:prep} since each problem has standalone type and function
definitions, making their separation trivial. The problem set includes known
examples from the software verification and inductive theorem proving
literature, including higher-order functions, common datatypes, etc., ensuring
that benchmark results are relevant to those fields. It also includes tools to
convert the problems into a variety of languages, including Haskell and
Isabelle.

TIP version 0.2 contains FIXME theorem statements, each of which is accompanied
by definitions of the types and functions it uses (other than those for integers
and booleans, which are ``built in''). There are FIXME function definitions, of
which most are duplicates. We remove $\alpha$-equivalent definitions in the
following way:

\begin{enumerate}
\item We prefix all type and function names with the filename in which we found
  the definition, so that e.g. the theorem statement from
  \texttt{tip2015/int\_add\_assoc.smt2} references the \texttt{plus} function from
  that file, rather than e.g. the \texttt{plus} function from
  \texttt{isaplanner/prop\_07.smt2}.
\item For each definition, all local variables (including function arguments,
  constructor parameters, pattern-matching variables and \texttt{let} bindings)
  are replaced by standard names \texttt{local-1}, \texttt{local-2}, etc. in a
  well-defined order.
\item An $O(n^2)$ comparison of each definition to every other definition is
  performed, to find any which are syntactically-identical modulo the names
  being defined (but \emph{not} the names being referenced, i.e. ). For those which are,
  we record which names are equivalent, replace all references to these names
  with whichever is smallest lexicographically, and remove
\end{enumerate}

Altogether TIP problem
by QuickSpec and IsaCoSy theory
exploration systems. To assess the scalability of their algorithms, we sampled
theories of various sizes: from theories containing only a single definition, up
to theories containing 20 definitions.
 We call the Theory Exploration Benchmark (TEB).

@@
